---
output:
  bookdown::pdf_book: # comment when rendering to word document
  # bookdown::word_document2: # uncomment if rendering to word document
    toc: no # change me to yes if rendering to Word
    keep_tex: no
    number_sections: yes
    fig_width: 6
    fig_height: 4
    latex_engine: xelatex
header-includes:
   - \usepackage{ragged2e}
   - \usepackage{setspace}
   - \usepackage{tocloft}
   - \usepackage{float} 
   - \floatplacement{figure}{H} 
   - \usepackage{wrapfig}
   - \usepackage{pdfpages}
   - \usepackage{caption}
bibliography: ./References/references.bib
csl: ./References/reference_format.csl
link-citations: yes
linkcolor: blue
---


```{r setup, include=FALSE, eval = T}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.duplicate.label = 'allow')

# These libraries are only needed for examples in the introduction 
library("png")
library("grid")
# These are needed in case you are having a glossary
library("knitr")
library("kableExtra")
library("dplyr")
library("tibble")
library("readr")
library(readxl)
```

<!-- Title page -->

```{r Cover, child = './Sections/00_Cover.Rmd', eval = T}
```

<!-- Set numbers in non Arabic format. And starts in page II as required by G+PS.
NOTE: Remove normal markdown chapters before rendering to PDF  -->
\pagenumbering{roman}
\setcounter{page}{2}

<!-- # Committee form -->

```{r Committee_form, child = './Sections/01_Committee_form.Rmd', eval = T}
```

\clearpage

<!-- Global page configuration -->

\setlength{\parindent}{4em} 
\linespread{1}
\doublespacing

<!-- # Abstract -->

\section*{Abstract}
\addcontentsline{toc}{section}{Abstract} 

```{r Abstract, child = './Sections/02_Abstract.Rmd', eval = T}
```

\clearpage

<!-- # Lay Summary -->

\section*{Lay Summary}
\addcontentsline{toc}{section}{Lay Summary}

```{r Lay_abstract, child = './Sections/03_Lay_abstract.Rmd', eval = T}
```

\clearpage

<!-- # Preface -->

\section*{Preface}
\addcontentsline{toc}{section}{Preface}

```{r Preface, child = './Sections/04_Preface.Rmd', eval = T}
```

\clearpage


<!-- Include a table of contents (including list of figures and tables) without numbering and with dots -->


\addcontentsline{toc}{section}{Table of Contents}

\thispagestyle{empty}
\begin{singlespace}
\renewcommand{\cftsecleader}{\cftdotfill{\cftdotsep}}
\setcounter{tocdepth}{3}
\tableofcontents
\clearpage

\addcontentsline{toc}{section}{List of Tables}
\listoftables
\clearpage

\addcontentsline{toc}{section}{List of Figures}
\listoffigures
\clearpage
\end{singlespace}


<!-- # Glossary -->
\section*{Glossary}
\addcontentsline{toc}{section}{Glossary}

```{r Glossary, child = './Sections/05_Glossary.Rmd', eval = T}
```

\clearpage

<!-- # Acknowledgements -->
\section*{Acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}

```{r Acknowledgements, child = './Sections/06_Acknowledgements.Rmd', eval = T}
```

\clearpage

<!-- Dedication -->
\section*{Dedication}
\addcontentsline{toc}{section}{Dedication}

```{r Dedication, child = './Sections/07_Dedication.Rmd', eval = T}

```


\clearpage

# Chapter One: Introduction

<!-- Set Numbering properly for rest of the document -->
\pagenumbering{arabic}

<!-- This code will needs to be before each chapter so figures and tables numbering are according to each chapter -->
<!-- Note that the chapter number changes each time -->

\renewcommand{\thefigure}{1.\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{1.\arabic{table}}
\setcounter{table}{0}
\renewcommand{\theequation}{1.\arabic{equation}}
\setcounter{equation}{0}

```{r Introduction, child = './Sections/08_Introduction.Rmd', eval = T}
```

\clearpage

<!-- From here on, include your different data chapters using normal Rmarkdown titles -->

<!-- Modify me! -->
# Chapter Two: Accurate ethnicity prediction from placental DNA methylation data

<!-- Set figure and table numbering according to the chapter -->
\renewcommand{\thefigure}{2.\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{2.\arabic{table}}
\setcounter{table}{0}
\renewcommand{\theequation}{2.\arabic{equation}}
\setcounter{equation}{0}

```{r chapter_two, child = './Sections/09_Chapter_2.Rmd', eval = T}
```

\clearpage

<!-- Modify me! -->
# Chapter Three: Cell-specific characterization of the placental methylome

<!-- Set figure and table numbering according to the chapter -->
\renewcommand{\thefigure}{3.\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{3.\arabic{table}}
\setcounter{table}{0}
\renewcommand{\theequation}{3.\arabic{equation}}
\setcounter{equation}{0}

```{r chapter_three, child = './Sections/10_Chapter_3.Rmd', eval = T}
```

\clearpage

<!-- Modify me! -->
# Chapter Four: Maternal contamination in placental DNA methylation studies

<!-- Set figure and table numbering according to the chapter -->
\renewcommand{\thefigure}{4.\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{4.\arabic{table}}
\setcounter{table}{0}
\renewcommand{\theequation}{4.\arabic{equation}}
\setcounter{equation}{0}

```{r chapter_four, child = './Sections/11_Chapter_4.Rmd', eval = T}
```

\clearpage

<!-- Done with data chapters, last sections -->

# Discussion

```{r Discussion, child = './Sections/12_Discussion.Rmd', eval = T}
```

\clearpage

<!-- # References -->
\section*{Bibliography}
\addcontentsline{toc}{section}{Bibliography} 

<!-- Reset page config. For proper referencing  -->
\noindent
\leftskip=2em
\parindent=-2em

<div id="refs"></div>

\clearpage

<!-- Go back to original page config. -->
\setlength{\parindent}{4em} 
\linespread{1}
\doublespacing


<!-- # Appendices -->
\section*{Appendices}
\addcontentsline{toc}{section}{Appendices} 

<!-- ## Appendix A -->

\section*{Appendix A - Supplementary information for "Name of Chapter Two"}
\addcontentsline{toc}{section}{Appendix A}

<!-- Set Figure, table and equations numbering according to the chapter -->
\renewcommand{\thefigure}{A2.\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{A2.\arabic{table}}
\setcounter{table}{0}
\renewcommand{\theequation}{A2.\arabic{equation}}
\setcounter{equation}{0}


```{r appendix_a, child = './Sections/13_Appendix_A.Rmd', eval = T}
```


<!-- ## Appendix B  -->

\section*{Appendix B - Supplementary information for "Name of Chapter Three"}
\addcontentsline{toc}{section}{Appendix B}

<!-- Set Figure, table and equations numbering according to the chapter -->
\renewcommand{\thefigure}{B3.\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{B3.\arabic{table}}
\setcounter{table}{0}
\renewcommand{\theequation}{B3.\arabic{equation}}
\setcounter{equation}{0}


```{r appendix_b, child = './Sections/14_Appendix_B.Rmd', eval = T}
```

<!-- ## Appendix C -->

\section*{Appendix C - Supplementary information for "Name of Chapter Four"}
\addcontentsline{toc}{section}{Appendix C}


<!-- Set Figure, table and equations numbering according to the chapter -->
\renewcommand{\thefigure}{C4.\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{C4.\arabic{table}}
\setcounter{table}{0}
\renewcommand{\theequation}{C4.\arabic{equation}}
\setcounter{equation}{0}


```{r appendix_c, child = './Sections/15_Appendix_C.Rmd', eval = T}
```


<!-- END OF SCRIPT -->
