---
output:
  bookdown::pdf_book:
    toc : no
    keep_tex: no
    number_sections: yes
bibliography: ../References/references.bib
---

(ref:figb1-cap) Fluorescence-activated cell-sorting and immune fluorescence staining. A) Fluorescence-activated cell-sorting (FACS) workflow schematic. B-E) Immunofluorescence staining (IF) of term cell-sorted sample with known characteristic cell type markers that were not selected for in the FACS procedure. Nuclei are shown via DAPI staining (blue). Scale bars: 100µm. B) Trophoblasts (KRT7: green, VIM: red). C) Hofbauer cells (CD68: green). D) Endothelial cells (CD31: green). E) Stromal cells (VIM: red).


```{r figb1, fig.cap="(ref:figb1-cap)", eval=TRUE, echo=T, echo=FALSE, fig.align='center', out.extra=''}
knitr::include_graphics(here::here('Images', 'appendix-b', 'Figure S1.pdf'))
```

<!----- Tables---->

\captionsetup{width=7in}

```{r tabb1, eval = FALSE, echo = T, echo=FALSE, warning = FALSE, message=FALSE,  out.extra=''}
readxl::read_csv(here::('Tables', 'appendix-b', 'Table S1.csv')) %>%
  kableExtra::kbl(
  booktabs = TRUE,
  caption = 'Title. Caption.')
```