---
output:
  bookdown::pdf_book:
    toc : no
    keep_tex: no
    number_sections: yes
bibliography: ../References/references.bib
---

(ref:figc1-cap) Samples that failed 1 or more 27 Illumina data quality control checks had higher P(outlier) than samples that passed quality control.


```{r figc1, fig.cap="(ref:figc1-cap)", fig.scap = "DNAm data quality and P(outlier)", eval=T, echo=T, echo=FALSE, fig.align='center', out.extra=''}
knitr::include_graphics(here::here('Images', 'appendix-c', 'Figure S1.pdf'))
```

(ref:figc2-cap) Principal component analysis (PCA) using placental and maternal tissue samples.

```{r figc2, fig.cap="(ref:figc2-cap)", fig.scap = "Principal component analysis using placental and maternal tissue samples", eval=T, echo=T, echo=FALSE, fig.align='center', out.extra=''}
knitr::include_graphics(here::here('Images', 'appendix-c', 'Figure S2.pdf'))
```

(ref:figc3-cap) Epigenetic age analysis. A) Adult epigenetic age (EA) estimated on placental samples was not associated with reported gestational age. B) Adult EA was associated with P(outlier) in 2 placental datasets. C) Placental epigenetic gestational age estimated in decidua, maternal blood, and chorionic villi samples was compared to reported gestational age.

```{r figc3, fig.cap="(ref:figc3-cap)", fig.scap = "Epigenetic age analysis", eval=T, echo=T, echo=FALSE, fig.align='center', out.extra=''}
knitr::include_graphics(here::here('Images', 'appendix-c', 'Figure S3.pdf'))
```

<!----- Tables---->

\captionsetup{width=7in}

```{r tabc1, eval = TRUE, echo = T, echo=FALSE, warning = FALSE, message=FALSE,  out.extra=''}
readxl::read_csv(here::here('Tables', 'appendix c', 'Table S1.csv')) %>%
  kableExtra::kbl(
  booktabs = TRUE,
  caption = 'Number of samples that failed quality control checks per dataset.')
```